HEP Software Foundation Community White Paper Working Group -- Data Organization, Management and Access (DOMA)

### 2018-04-16 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
CWP DOMA: import fixes from Overleaf
- Changes made up to 13/4 included

\abstract{Without significant changes to data organization, management, and access (DOMA), HEP experiments will find scientific output limited by how fast data can be accessed and digested by computational resources. In this white paper we discuss challenges in DOMA that HEP experiments such as the HL-LHC will face as well as potential ways to address them. A research and development timeline to assess these changes is also proposed.}

@@

\title{HSF Community White Paper on Data Organization, Management and Access
(DOMA)}

@@

The coming decade will bring HEP to the exabyte (EB) scale with the expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as the Square Kilometre Array (SKA) entering this
regime. In devising experimental computing models for this era, many
factors have to be taken into account. In particular, the increasing
availability of very high-speed networks, which may reduce the need for
% * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
%
% > very high-speed networks
% What is "very high-speed" here? 100G? 1000G?
%
% ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
CPU and data co-location, need to be examined. Such networks may allow
for more extensive use of data access over the wide-area network (WAN),
which may provide fail-over capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation and analysis models, such as a potential move to or
sub-event-based or event-based data streaming from the more
% * <briedel@uchicago.edu> 2018-04-04T18:00:17.686Z:
%
% > sub-event-based or event-based data
% Added "sub-event-based" because we talk about it below.
%
% ^ <briedel@uchicago.edu> 2018-04-04T18:00:36.656Z.
traditional dataset-based or file-based data access, will be
particularly important for optimizing the utilization of
opportunistic computing cycles on HPC facilities, commercial cloud resources,
and campus clusters, and can potentially resolve currently
limiting factors such as job eviction.
% * <briedel@uchicago.edu> 2018-04-04T18:01:55.490Z:
%
% >  job eviction
% Using "job evictions" as a reason to move to different DOMA model seems like an odd choice. Shouldn't it be something like "scheduling difficulties"
%
% ^.

Planned analysis of HL-LHC data and data analysis from other large
% * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
%
% > large
% XENON1T has shown that the collaboration doesnt have to be "large" to require DOMA
%
% ^.
international collaborations will need to adopt a distributed computing
model due to its scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

@@

The LHC experiments currently collectively provision and manage about an EB of
storage of which approximately half is archival and half is
traditional disk storage. The annual storage requirements are expected
to jump by a factor of 10 for the HL-LHC. This increase is faster than
projected Moore's Law gains and will present major challenges. Storage
% * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
%
% > Moore's Law gains
% This is only a CPU. What about storage technology? Currently we are adding about 2 TB/year in disk capacity. It is unlikely will be at 100 TB/disk in 2023. We may have O(10) - O(100) TB SSDs  at a reasonable cost, but we can't be certain. What does the future of tape look like?
%
% ^.
will remain one of the visible cost drivers for HEP computing with
the projected growth and cost of the computational resources needed to
analyze the data expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

The three main challenges for data in the HL-LHC era can thus be
summarized:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item \textbf{Big Data:} The HL-LHC era will significantly increase both expected
  date rates and data volumes. Computing systems will need to
  handle this without significant cost increases and within evolving
  storage technology limitations.
\item \textbf{Dynamic Distributed Computing:} In addition, the significantly
  increased computational requirements for the HL-LHC era will also
  place new requirements on data. Specifically, the use of new types of
  compute resources (e.g., cloud, HPC) with different dynamic
  availability and characteristics will require more dynamic
  DOMA systems.
\item \textbf{New Applications:} New applications, such as machine learning
  training or high-rate data-query systems for analysis, will likely be
  employed to meet the computational constraints and to extend the
  physics reach of the HL-LHC. These new applications will place new
  requirements on how and where data is accessed and produced. For
  example, specific applications (e.g., training for machine learning)
  may require use of specialized processor resources, such as GPUs,
% * <briedel@uchicago.edu> 2018-04-04T18:14:36.074Z:
%
% > ,  placing further requirements on data.
% Not sure this is needed.
%
% ^.
  placing further requirements on data.
\end{enumerate}

The rapid increase in recent years of data-intensive problems in both
the commercial world and in the rest of the research world also provides
a number of opportunities and solution to tackle these challenges.

@@

\begin{itemize}
\item \textbf{Data Organization:} This is essentially how data is structured
  as it is written. Most data is written in flat files, in the ROOT format, and
  typically with a column-wise organization of the data. The records
  corresponding to these columns are compressed. The internal details of
  this organization are typically visible only to individual software
  applications.
\item \textbf{Data Management:} The key challenge here was the transition to
  the use of distributed computing in the form of the grid. The
  experiments developed dedicated data catalog,  transfer and placement systems,
  to move data between computing centers. To first
  order, the computing models were rather static: data was placed at
  sites and the relevant compute jobs were sent to the right locations.
  Applications might interact with catalogs or, at times, the workflow
  management systems does this on behalf of the applications.
\item \textbf{Data Access:} Various protocols are used for direct reads
  (rfio, dcap, xrootd, etc.) with a given computer center and/or
  explicit local stage-in and caching for read by jobs. Application
  access may use different protocols than those used by the data
  transfers between sites.
\end{itemize}

Before the LHC turn-on and in the first years of data taking, these three
areas were, to first order, optimized independently. Many of the
challenges were in the area of ''Data Management (DM)'' as the Worldwide
LHC Computing Grid was commissioned. As the LHC experiments' computing matured
through Run1 and Run2, interest has turned to optimizations spanning
these three areas. For example, the recent use of ``Data Federations''
must consider together both Data Management and Access. As we will see below, some of the
foreseen opportunities towards HL-LHC may require global optimizations.

Thus, in this document we take a broader view than traditional ``DM'',
and consider the combination of ``Data Organization, Management and
Access (DOMA)'' together. We believe that this full picture of data
needs in HEP will provide important opportunities for efficiency and
scalability as we enter the many-EB era.

@@

The way data are used at HEP experiments results in a dynamic range that
spans many orders of magnitude between the IOPS transactions for data
access and the bandwidth to storage that is required to store or
retrieve the information. The primary data analysis workflows that drive
these bandwidth requirements can be categorized into four main
activities. These are the contextual environment in which we present
future evolutions for HEP data access. In particular, characteristics
that describe these activities are enumerated for each category:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
  the complexity of detector data and the computational algorithms
  needed to associate and disentangle the data. This is true of
  reconstruction methods used today at the LHC and throughout the HEP
  community. This will continue to be the driving characteristic of
  reconstruction at the HL-LHC and for liquid argon TPC experiments. The
  CPU-bound nature of the algorithms and the ratio of event size to
  network bandwidth make the option of streaming individual events
  records (\emph{e.g.} full collision events or beam spill triggers)
  across wide area networks to compute elements a feasible analysis
  strategy. The requirements of event streaming analysis models can be
  satisfied by a wide range of storage solutions.
\item \textbf{Simulation:} HEP event simulation strategies focus on
  transforming a small set of key input parameters into event and
  detector response information that is representative of what would be
  seen in the physical detector systems. Simulation processes are
  typically characterized by highly asymmetric Input/Output workflows
  which have very small data ingest (typically just the key input
  parameters) and produce output events which are similar in size and
  structure to real detector data. The output of these simulation
  processes overlap heavily with other data storage and data replication
  processes. The requirements and challenges that this output presents
  are treated in those contexts. The primary drivers that differentiate
  this type of HEP workflow from the other categorizations are the
  increased complexities that the simulation stages often require. These
  simulations often rely on auxiliary event generation mechanisms and
  can, in some HEP domains, require large external data sets
  representing interaction cross sections or detector response functions
  to be available to each instance of the event simulation. This type of
  common input and overlay data may present a challenge in relation to
  its scaling to the runtime environments of future HPC facilities and
  to the design of caching layers at computing site.
\item \textbf{Analysis:} Analysis level datasets and subsets are typically
  comprised of refined or reduced information that is most pertinent to
  the extraction of physics quantities. These analysis-focused event
  records can often be accepted or rejected through fast pre-filtered
  methods which do not require the reading and retrieval of the full
  event record. This fundamental difference in the way that the analysis
  data is consumed makes it more susceptible to storage bandwidth
  limitations and to IOPs transaction limits that underlying
  technologies can have. In particular, the internal structure of the
  event data can dramatically impact the efficiency of data retrieval
  and favor different access models and protocols (i.e. event
  streaming) while the actual analysis calculations being performed can
  dramatically affect the CPU-to-IO ratios. These factors make this
  categorization of data the most challenging to project to future
  storage needs and models.
\item \textbf{Replication:} Data replication, whether within a site or
  across multiple sites, is performed to enable efficient data access by
  improving data locality (the proximity of the data to compute
  resources) and exploiting the available data bandwidth between the
  storage and the compute locations. Data replication also serves to
  guard against data loss. Data replication places modest requirements
  on the rate of IOP transactions that need to be supported, but drives
  the bandwidth requirements that are needed in enacting site-to-site
  data transfers between WAN endpoints (e.g. 10-100~Gbps). This
  requirement on the available WAN bandwidth provides a corresponding
  performance requirement on on the underlying storage systems to match
  network performance. Satisfying both requirements is necessary to
  provide efficient point-to-point data flow.
\end{enumerate}

In addition to the characteristics described above, HEP workflows in
each category exhibit a high degree of parallelism, which allow the work
to map readily into both high throughput computing (HTC) and high
performance computing (HPC) environments. These domains span wide
dynamic ranges in their requirements for concurrent data access. If the
experiments of the HL-LHC, DUNE and other large international projects
are to utilize these facilities, they will need the ability to integrate
with multiple storage technologies. The storage technologies will need
to be tuned or matched to the corresponding computing resources that the
experiments are targeting.

These storage models will also need to be designed and optimized for
cost, in the context of the scientific workflows that they are enabling.
Simultaneously, these storage models must remain flexible enough to
support evolving physical storage options similar to the the shifts that
have occurred in recent years between mRAM, SSDs, shingled disks, tapes,
and other technologies.

@@

The HEP community has long-established data organization patterns that have
shaped the prevailing data access and storage
models that are in use today. These established patterns are centered
around sets of event records organized into discrete collections through
traditional file-based storage. There are, however, potential gains that
can be made by exploring alternative data organization models and the
corresponding access and storage models that they can enable. These
alternative designs should be evaluated for the HL-LHC era and, if found
to be advantageous, implemented by the community. A shift in designs may
have the potential to reduce storage costs, shift network utilization,
and result in more efficient use of compute resources.

@@

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (\emph{e.g.}, physics
object). Any such catalogs may have to be external to the primary data
store (\emph{e.g.}, relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
% * <briedel@uchicago.edu> 2018-04-05T02:04:26.292Z:
%
% > content-addressable
% This comes a little out of nowhere. Where is this discussed before?
%
% ^.
handle to have a fully descriptive location of a given event or other
object being sought.
% * <briedel@uchicago.edu> 2018-04-05T02:05:03.704Z:
%
% > or other object
% This is the first time "other object" is mentioned. What other object? TBranches? It would be better to call this sub-event data like done before.
%
% ^.

@@

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
% * <briedel@uchicago.edu> 2018-04-05T02:08:36.870Z:
%
% > Akamai.
% Google also offers one
%
% ^.
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance where a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products that minimizes data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role. NDNs allow the naming, at the
network level, of the data you are looking for rather than its location, which
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, ARC-style caching) and may
% * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
%
% > server-less caching
% In-memory caching?
%
% ^.
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

@@

For more scalable analysis, another opportunity to evaluate is how much
work can be offloaded to a storage system, for example caching
uncompressed or reordered data for fast access. The idea can be extended
to virtual data and to query interfaces which would perform some of the
transformation logic currently executed on CPU workers. Interactive
querying of large datasets is an active field in the Big Data industry;
examples include Spark-SQL, Impala, Kudu, Hawq, Apache Drill, and Google
Dremel/BigQuery. A key question is about the usability of these
techniques in HEP and we need to assess if our data transformations are
not too complex for the SQL-based query languages used by these
products. We also need to take into account that the adoption of these
techniques, if they prove to be beneficial, would represent a disruptive
change which directly impacts the end user and therefore promoting
acceptance through intermediate solutions would be desirable.

@@

As the size of available datasets grows, we must allow for the
possibility of increased user-driven selection and dataset generation.
In terms of reducing the strain on data access mechanisms, techniques such as data
augmentation can prove useful. Cluster-computing frameworks (e.g.,
Spark) can also be leveraged to reduce the time of dataset reduction and
access. It will be necessary to study and classify I/O patterns in
applications used in order to understand how data access methods can be
optimized across prevalent access patterns. Pattern data can then be
used in algorithms resulting from active research in computer science
that study trade-offs between storage and CPU in various computing models.

@@

HEP experiments should be prepared to leverage ``tactical storage''.

Storage that becomes most cost-effective as it becomes available (e.g.,
from a cloud provider) and have a data management and provisioning
system that can exploit such resources on short notice. Volatile data
sources would impact many aspects of the system; catalogs, job
brokering, monitoring/alerting, accounting, the applications themselves.

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance trade-offs.
Currently, tape is extensively used to hold data that cannot economically be made available online. While the tape-stored data is still accessible,
it comes with a high latency penalty and thus limiting possible analysis. We suggest investigating either separate direct-access-based archives (e.g.
disk or optical) or new models that overlay online direct-access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.

Cost reductions in maintenance and operation of the storage
infrastructure can be realized through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a monoculture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialized systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavors that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

@@

\begin{itemize}
\item Quantify impact on performances
  and resource utilization (storage, network) for the main type of
  access patterns ({\it i.e.} simulation, reconstruction, analysis).
\item Assess the impact on catalogs and
  data distribution.
\item Assess whether
  event-granularity makes sense in object stores that tend to require
  large chunks of data for efficiency.
\item Test for improvement in recoverability
  from preemption, in particular when using cloud spot resources and/or
  dynamic HPC resources.
\end{itemize}

\emph{The previous tasks should be completed by 2020 (and can be performed in parallel).}

@@

\begin{itemize}
\item Evaluate row-based versus column-based
  data organization on the performance of each type of access used.
\item Investigate data storage and access solutions that support the use of map-reduce or
Spark-like analysis tools and their adaptability to HEP analysis needs.
\item Evaluate new compression schemes, just-in-time
  decompression schemes and mappings onto hardware architectures
  considering the flow of data from spinning disk to memory and
  application.
\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filterting through metdata
\end{itemize}

\emph{The proposed proof-of-concept involves the previous tasks that should be completed by 2020.
If successful, implementation as a production system will be done in the following years.}

@@

-\emph{The first two actions should be completed by 2020 but CDN/NDN
-evaluation is more long-term}

>>>>

\abstract{Without significant changes to data organization, management, and
access (DOMA), HEP experiments will find scientific output limited by how
fast data can be accessed and digested by computational resources. In this
white paper we discuss challenges in DOMA that HEP experiments, such as
the HL-LHC, will face as well as potential ways to address them. A research
and development timeline to assess these changes is also proposed.}

@@

\title{HEP Software Foundation Community White Paper Working Group --
Data Organization, Management and Access (DOMA)}

@@

The coming decade will bring HEP to the exabyte scale with the expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as the Square Kilometer Array (SKA) entering this
regime. In devising computing models for this era, many
factors have to be taken into account. In particular, the increasing
availability of very high-speed networks may reduce the need for
% * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
%
% > very high-speed networks
% What is "very high-speed" here? 100G? 1000G?
%
% ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
CPU and data co-location. Such networks may allow
for more extensive use of data access over the wide-area network (WAN) while providing fail-over capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation and analysis models, such as a potential move to event-based or
sub-event-based data streaming from the more traditional dataset-based or file-based
data access, will be particularly important for optimizing the
utilization of opportunistic computing cycles on HPC facilities,
commercial cloud resources, and campus clusters, and can potentially
resolve currently limiting factors such as job eviction.

Planned analysis of HL-LHC data and data analysis from other
% * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
%
% > large
% XENON1T has shown that the collaboration doesnt have to be "large" to require DOMA
%
% ^ <briedel@uchicago.edu> 2018-04-09T20:36:43.846Z.
international collaborations will need to adopt a distributed computing
model due to its scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

@@

The LHC experiments currently collectively provision and manage about an Exabyte of
storage of which approximately half is archival and half is
traditional disk storage. The annual storage requirements are expected
to jump by a factor of 10 for the HL-LHC. This increase is faster than
projected density gains in current storage technologies and thus will present major challenges. Storage
% * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
%
% > Moore's Law gains
% This is only a CPU. What about storage technology? Currently we are adding about 2 TB/year in disk capacity. It is unlikely will be at 100 TB/disk in 2023. We may have O(10) - O(100) TB SSDs  at a reasonable cost, but we can't be certain. What does the future of tape look like?
% changed to "density gains in current storage technologies." Tape scaling has quite a future but hard drives, less so
% ^.
will remain one of the visible cost drivers for HEP computing. The projected growth and cost of the computational resources needed to
analyze the data is expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

The three main challenges for data in the HL-LHC era can be
summarized as:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item \textbf{Big Data:} The expected data volume will significantly increase in the HL-LHC era. Computing systems will need to
  handle this without significant cost increases and within evolving
  storage technology limitations.
\item \textbf{Dynamic Distributed Computing:} In addition, the significantly
  increased computational requirements for the HL-LHC era will also
  place new requirements on data. Specifically, the use of new types of
  compute resources (e.g., cloud, HPC) with different dynamic %don't want this to be just cloud and HPC, hence "new types"
  availability and characteristics will require more dynamic
  DOMA systems.
\item \textbf{New Applications:} New applications, such as machine learning
  training or high-rate data-query systems for analysis, will likely be
  employed to meet the computational constraints and to extend the
  physics reach of the HL-LHC. These new applications will place new
  requirements on how and where data is accessed and produced. For
  example, specific applications (e.g., training for machine learning)
  may require use of specialized processor resources, such as GPUs.
\end{enumerate}

The rapid increase in recent years of data-intensive problems in both
the commercial world and in the rest of the research world also provides
a number of opportunities and solutions to tackle these challenges.

@@

\begin{itemize}
\item \textbf{Data Organization:} This is essentially how data is structured
  as it is written. Most data is written in the ROOT file format. ROOT serializes data objects, compresses, and writes them in a column-wise manner. The data written in this way can only be meaningfully read  by a software that includes data object libraries.
\item \textbf{Data Management:} The key challenge here was the transition to
  the use of distributed computing in the form of the grid. To tackle this, the
  experiments developed dedicated data catalog,  transfer and placement systems. To first
  order, the computing models were rather static: data was placed at
  sites and the relevant compute jobs were sent to the right locations.
\item \textbf{Data Access:} Various protocols are used for direct reads
  (RFIO, dCap, XRootD, etc.) within a given computer center and/or
  explicit local stage-in and caching for read by jobs. Application
  access may use different protocols than those used by the data
  transfers between sites.
\end{itemize}

Before the LHC turn-on and in the first years of data taking, these three
areas were, to first order, optimized independently. Many of the
challenges were in the area of ``Data Management'' (DM) as the Worldwide
LHC Computing Grid was commissioned. As the LHC experiments' computing matured
through Run1 and Run2, interest has turned to optimizations spanning
these three areas. For example, the recent use of ``Data Federations''
must consider together both Data Management and Access. As we will see below, some of the
foreseen opportunities towards HL-LHC may require global optimizations.

Thus, in this document we take a broader view than traditional DM,
and consider the combination of ``Data Organization, Management and
Access '' together. We believe that this full picture of data
needs in HEP will provide important opportunities for efficiency and
scalability as we enter the many-Exabyte era.

@@

HEP data transfer and access patterns require network and storage architectures to be able to sustain a large dynamic range in IOPS. The primary data analysis workflows that drive
these bandwidth requirements can be categorized into four main
activities:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item \textbf{Reconstruction:} Event reconstruction is CPU limited, due to
  the complexity of detector data and the computational algorithms
  needed to associate and disentangle the data. This is true of
  reconstruction methods used today at the LHC and throughout the HEP
  community and will continue to be the driving characteristic of
  reconstruction at the HL-LHC and for liquid argon TPC experiments. The
  CPU-bound nature of the algorithms and the ratio of event size to
  network bandwidth make the option of streaming individual events
  records (e.g., full collision events or beam spill triggers)
  across the WAN to compute elements a feasible
  strategy. The requirements of event streaming can be
  satisfied by a wide range of storage solutions.
\item \textbf{Simulation:} HEP event simulation strategies focus on
  transforming a small set of key input parameters into event and
  detector response information that is representative of what would be
  seen in the physical detector systems. Simulation processes are
  typically characterized by highly asymmetric IO. They typically have very small data ingest (typically just the key input
  parameters), while output data are similar in size and
  structure to real detector data. Due to their similar characteristics, output data from simulation processes share the requirements and associated challenges of detector data. The primary drivers that differentiate
  this type of HEP workflow from the other categorizations are the
  increased complexities that the simulation stages often require. These
  simulations often rely on auxiliary event generation mechanisms and
  can, in some HEP domains, require large external data sets
  representing interaction cross sections or detector response functions
  to be available to each instance of the event simulation. This type of
  common input and overlay data may present a challenge in relation to
  its scaling to the runtime environments of future HPC facilities and
  to the design of caching layers at computing site.
\item \textbf{Analysis:} Analysis level datasets and subsets are typically
  comprised of refined or reduced information that is most pertinent to
  the extraction of physics quantities. These analysis-focused event
  records can often be accepted or rejected through fast pre-filtered
  methods which do not require the reading and retrieval of the full
  event record. This fundamental difference in the way that the analysis
  data is consumed makes it more susceptible to storage bandwidth
  limitations and to IOPS transaction limits that underlying
  technologies can have. In particular, the internal structure of the
  event data can dramatically impact the efficiency of data retrieval
  and favor different access models and protocols while the actual analysis calculations being performed can
  dramatically affect the CPU-to-IO ratios. These factors make this
  categorization of data the most challenging to project to future
  storage needs and models.
\item \textbf{Replication:} Data replication, whether within a site or
  across multiple sites, is performed to enable efficient data access by
  improving data locality (the proximity of the data to compute
  resources) and exploiting the available data bandwidth between the
  storage and the compute locations. Data replication also serves to
  guard against data loss. Data replication places modest requirements
  on the IOPS that need to be supported, but drives
  the bandwidth requirements that are needed in enacting site-to-site
  data transfers between WAN endpoints (\emph{e.g.}, 10-100~Gbps). This
  requirement on the available WAN bandwidth provides a corresponding
  performance requirement on on the underlying storage systems to match
  network performance. Satisfying both requirements is necessary to
  provide efficient point-to-point data flow.
\end{enumerate}

In addition to the characteristics described above, HEP workflows in
each category exhibit a high degree of parallelism, which allow the work
to map readily into both high throughput computing (HTC) and HPC environments. These domains span wide
dynamic ranges in their requirements for concurrent data access. If the
experiments of the HL-LHC, DUNE and other international projects
are to utilize these facilities, they will need the ability to integrate
with multiple storage technologies. The storage technologies will need
to be tuned or matched to the corresponding computing resources that the
experiments are targeting.

These storage models will also need to be designed and optimized for
cost, in the context of the scientific workflows that they are enabling.
Simultaneously, these storage models must remain flexible enough to
support evolving physical storage options similar to the the shifts that
have occurred in recent years between NVMe, SSDs, shingled disks, tapes,
and other technologies.

@@

The HEP community has long-established data organization patterns that have
shaped the prevailing data access and storage
models that are in use today. Today's HEP data organization is file-centric with sets of events stored in a given file. Alternative data organization models could yield gains depending on their respective data access and storage models. These
alternative designs should be evaluated for the HL-LHC era, as they may produce cost reduction, better network utilization, and efficiency gains.

@@

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (e.g. physics
object). Any such catalogs may have to be external to the primary data
store (e.g., relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
% * <briedel@uchicago.edu> 2018-04-05T02:04:26.292Z:
%
% > content-addressable
% This comes a little out of nowhere. Where is this discussed before?
%
% ^.
handle to have a fully descriptive location of a given event or other
object being sought.
% * <briedel@uchicago.edu> 2018-04-05T02:05:03.704Z:
%
% > or other object
% This is the first time "other object" is mentioned. What other object? TBranches? It would be better to call this sub-event data like done before.
%
% ^.

@@

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
% * <briedel@uchicago.edu> 2018-04-05T02:08:36.870Z:
%
% > Akamai.
% Google also offers one
%
% ^.
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance where a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products, minimizing data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role. NDNs allow the naming, at the
network level, of the data you are looking for rather than its location, which
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, Adaptive Replacement Cache-style caching) and may
% * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
%
% > server-less caching
% In-memory caching?
%
% ^.
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

@@

For more scalable analysis, another opportunity to evaluate is how much
work can be offloaded to a storage system, for example caching
uncompressed or reordered data for fast access. The idea can be extended
to virtual data and to query interfaces which would perform some of the
transformation logic currently executed on CPU workers. Interactive
querying of large datasets is an active field in the Big Data industry;
examples include Spark-SQL, Impala, Kudu, Hawq, Apache Drill, and Google
Dremel/BigQuery. A key question is about the usability of these
techniques in HEP and we need to assess if our data transformations are
not too complex for the SQL-based query languages used by these
products. We also need to take into account that the adoption of these
techniques, if they prove to be beneficial, would represent a disruptive
change which directly impacts the end user and therefore promoting
acceptance through intermediate solutions would be desirable. One such solution - ``Service X'' - has been proposed in \href{https://bit.ly/2Ix9NIL}{Organizing Data Lakes}. The service would start with the data in its persistent format, perform one or more operations on it  (\emph{e.g.} joins, query, decompression, filtering to sub-event level), and finally deliver a data stream to a client.

@@

As the size of available datasets grows, we must allow for the
possibility of increased user-driven selection and dataset generation.
In terms of reducing the strain on data access mechanisms, techniques such as data
augmentation can prove useful. Cluster-computing frameworks (e.g.,
Spark) can also be leveraged to reduce the time of dataset reduction and
access. It will be necessary to study and classify I/O patterns in
applications used in order to understand how data access methods can be
optimized across prevalent access patterns. Pattern data can then be
used in algorithms resulting from active research in computer science
that studies trade-offs between storage and CPU in various computing models.

@@

HEP experiments should be prepared to leverage ``tactical storage''.
Such storage may be provisioned only when it becomes cost-effective
(e.g.,
% * <briedel@uchicago.edu> 2018-04-05T15:43:01.526Z:
%
% > most
% "more" rather than "most"?
%
% ^.
from a cloud provider) and have a data management and provisioning
system that can exploit such resources on short notice. Volatile data
% * <briedel@uchicago.edu> 2018-04-05T15:43:37.726Z:
%
% > Volatile data sources would impact many aspects of the system; catalogs, job brokering, monitoring/alerting, accounting, the applications themselves.
% Aren't we dealing with somewhat volatile storage in terms of xrootd caches today already? The file _might_ be there, but it isnt guaranteed.
%
% ^.
sources would impact many aspects of the system; catalogs, job
brokering, monitoring/alerting, accounting, the applications themselves.

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance trade-offs.
Currently, tape is extensively used to hold data that cannot economically be made available online. While the tape-stored data is still accessible,
it comes with a high latency penalty and thus limiting possible analysis. We suggest investigating either separate direct-access-based archives (e.g.
disk or optical) or new models that overlay online direct-access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.
% * <briedel@uchicago.edu> 2018-04-05T15:49:15.646Z:
%
% > stability.
% This paragraph begs the question of whether one needs to move to "direct-access-based archives" if the data is only accessed once or twice. The plenty of tape is not that great in that case.
%
% ^.

Cost reductions in maintenance and operation of the storage
infrastructure can be realized through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a mono-culture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialized systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavors that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

@@

\begin{itemize}
\item Quantify impact on performance
  and resource utilization (storage, network) for the main type of
  access patterns ({\it i.e.} simulation, reconstruction, analysis).
\item Assess the impact on catalogs and
  data distribution.
\item Assess whether
  event-granularity makes sense in object stores that tend to require
  large chunks of data for efficiency.
\item Test for improvement in recoverability
  from preemption, in particular when using cloud spot resources and/or
  dynamic HPC resources.
\end{itemize}

\emph{The above tasks should be completed by 2020 (and can be performed in parallel).}

@@

\begin{itemize}
\item Evaluate row-based versus column-based
  data organization on the performance of each type of access used.
\item Investigate data storage and access solutions that support the use of map-reduce or
Spark-like analysis tools and their adaptability to HEP analysis needs.
\item Evaluate new compression schemes, just-in-time
  decompression schemes and mappings onto hardware architectures
  considering the flow of data from spinning disk to memory and
  application.
\item Evaluate possible alternative storage formats to the ROOT format, esp.\ storage formats optimized for storage efficiency and/or  data filtering through metadata
\end{itemize}

\emph{The proposed proof-of-concept involves the above tasks that should be completed by 2020.
If successful, implementation as a production system will be done in the following years.}

@@

-\emph{The first two actions should be completed by 2020 but evaluation of CDN/NDN is more long-term}
